\startarticle 
\shrtitle{Detecting At-Risk Students...}
\shrauthor{E.~Rodriguez, O.~Kosheleva, V.~Kreinovich}{E.~Rodriguez, O.~Kosheleva, V.~Kreinovich}
\udc{378:219.2} 
\selectlanguage{english}
\date{02.12.2018}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Russian. If not compiled, it replaces the Russian text to English %%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Выявление подверженных риску студентов: эмпирические результаты и их теоретическое объяснение} 
\author{Э.Д.Р. Веласкес}{докторант}{edrodriguezvelasquez@miners.utep.edu}
\author{О. Кошелева}{к.ф.-м.н., доцент}{olgak@utep.edu}
\author{В. Крейнович}{к.ф.-м.н., профессор}{vladik@utep.edu}
\affil{Техасский университет в Эль Пасо, США} 
\abstract{При обучении очень важно как можно раньше выявлять студентов, которые могут быть подвержены риску провала. 
Традиционно для этой идентификации используются два естественных критерия: плохие оценки за предыдущие 
дисциплины и плохие оценки за первые задания в текущей дисциплине. Наши эмпирические результаты показывают, что 
эти критерии не всегда работают: иногда учащиеся, которые считаются подверженными риску по одному из этих критериев, 
стабильно преуспевают, а иногда учащийся, который не считается подверженным риску, часто терпит неудачу. 
В этой статье мы даем теоретическое объяснение наших количественных эмпирических результатов, и мы используем эти результаты, 
чтобы предоставить рекомендации о том, как лучше выявлять учащихся из группы риска.}
\keywords{обучение, выявление подверженных риску студентов, прогнозирование оценок учащихся}

\titleEng{Detecting At-Risk Students: Empirical Results and Their Theoretical Explanation}     
\authorEng{Edgar Daniel Rodriguez Velasquez}{Doctoral Student}{edrodriguezvelasquez@miners.utep.edu}
\authorEng{Olga Kosheleva}{Ph.D. (Phys.-Math.), Associate Professor}{olgak@utep.edu}
\authorEng{Vladik Kreinovich}{Ph.D. (Phys.-Math.), Professor}{vladik@utep.edu}
\affilEng{University of Texas at El Paso, El Paso, Texas 79968, USA} 
\abstractEng{In teaching, it is very important to identify,
as early as possible, students who may be at risk of failure.
Traditionally, two natural criteria are used for this
identification: poor grades in previous classes, and poor grades
on the first assignments in the current class. Our empirical
results show that these criteria do not always work: sometimes a
students deemed at-risk by one of these criteria consistently
succeeds, and sometimes a student who is not considered at-risk
frequently fails. In this paper, we provide a theoretical
explanation of our quantitative empirical results, and we use
these results to provide recommendations on how to better detect
at-risk students.} 
\keywordsEng{teaching, detecting at-risk students, predicting student grades}

\maketitle

\section{Formulation of the Problem}

\noindent{\bf How do we identify students that need more
attention?} Instructors and teaching assistants have a limited
amount of time to attend to (often a large number of) students. To
spend this time wisely, it is important to identify at-risk
students, students who need additional attention to succeed. How
can we identify these students?
\medskip

\noindent{\bf How do we identify at-risk students before class
starts: a natural idea.} Before the class starts, the only
information that we have to identify at-risk students is their
past performance. A good indication of the student's past
performance is their average grade in previous classes (in the US,
this average grade is called Grade Point Average, GPA for short).
If this GPA is low, close to the failure level, it is reasonable
to assume that the student is at risk of failure, and additional
efforts need to be taken to help these students.
\medskip

\noindent{\bf Once the class started, what additional information
can we use?} Once the course starts, we get the average grade from
the first few assignments --- and again, these average grades can
serve as another indicator of at-risk students: \begin{itemz}
\item if a student has been doing well so far, probably he or she
does not need special attention; \item on the other hand, if the
student's grades so far on the course's assignments, quizzes, and
tests were low, this seems like a good indication that a student
needs extra help.
\end{itemz}
\medskip

\noindent{\bf But is this indeed a reasonable strategy?} This may
sound reasonable, but it is a good idea to check if this strategy
indeed works.
\medskip

\noindent{\bf What we do in this paper.} In our previous research
\cite{Rodriguez2017}, we provided an empirical analysis of the
above strategies. In this paper, we recall these results, describe
a possible theoretical justification for these empirical results,
and provide resulting pedagogical recommendations.

\section{Empirical Results: Reminder}

\noindent{\bf What we did.} In \cite{Rodriguez2017}, we studied
student performance in several classes, including their starting
grade and their final grade in each class.
\medskip

\noindent{\bf What we expected.} We expected that in most of the
cases, both GPA in the previous classes and the average grade for
first few assignments would be very good predictors for student's
success.

This would mean that all our attention should be concentrated on
students with low GPA and/or low performance in the first
assignment, and we should not worry that much about other
students.
\medskip

\noindent{\bf What we found out: somewhat disappointing news.}
What we found out is that, surprisingly, overall, neither of two
usual criteria is a good predictor of the student's success:
\begin{itemz}
\item the correlation between the student's success and GPA is
low, and \item the correlation between the student's success in a
class and the student's average grade for the first few class
assignment is also very low.
\end{itemz}

At first glance, this sounds pessimistic: no way that we can
predict the student's success with good accuracy and so, we cannot
simply dismiss students who perform well so far as not needing our
attention: they may be at risk as well.
\medskip

\noindent{\bf What we found out: interesting news.} However, a
deeper analysis revealed an interesting phenomenon with respect to
both correlations.

It turns out that for about a third of all students, their past
GPA is a good predictor of the grade in the class. In other words,
such students show steady performance, with a low standard
deviation of grades from the average. \begin{itemz} \item Some
of these students are straight-A students, maybe with a few
B's.\item Some are straight-B students, usually with a few As and
Cs.\item Some are straight-C students, with a few Bs and Ds. \item
Some are straight-D students, with most grades Ds and
fails.\end{itemz} For such students, predicting their success is
easy: \begin{itemz} \item if the student's GPA was passing, we
can be reasonably sure that the students will pass this class;
\item otherwise, if the student's previous GPA was close to
failing, then this student is clearly at-risk, and an additional
attention needs to be paid to these students.\end{itemz}

Thus is about the third of all students --- students with steady
performance. For the remaining two thirds of the students, there
is no correlation between their GPA and grade in the class.

Similarly, with respect to relation between the preliminary grade
in the class and final grade for the same class, there is a
similar division:
\begin{itemz}
\item For about a third of the students, their initial performance
in a class is a good predictor of the student's final grade. \item
However, for the remaining two thirds of the students, based on
the initial performance, we cannot meaningfully predict their
final grade.
\end{itemz}

Based on two criteria, we have two classes of ``predictable''
students: \begin{itemz} \item students whose success in the
class can be predicted based on their GPAs, and \item students
whose success in the class can be predicted based on their
performance in the first weeks of the class. \end{itemz} Each
group contains about a third of all the students. These two groups
intersect: for some students, their performance in the class can
be predicted {\it both} based on GPA and on their initial
performance in the class.

Overall about half of the students are, in this sense,
predictable. For the other half, we cannot predict the student's
success.

\section{Theoretical Explanation of the Empirical Results}

\noindent{\bf Why 1/3: analysis of the problem and the resulting
justification.} In general, there is a low positive correlation
between the final grade for the class and the average grade on the
first few assignments. What is a natural way to describe this
``low'' in precise numerical terms?

Intuitively, the use the word ``low'' means that we consider two
possible values of the positive correlation: low and high. We
would like to assign numerical values $\ell$ and $h$ to these
words. The only limitation is that $0<\ell<h<1$. Out of many pairs
$(\ell,h)$ that satisfy this inequality, which pair should we
choose?

We have no reason to believe that some such pairs are more
probable than others. Thus, it is reasonable to assume that all
these pairs are equally probable, i.e., that we have a uniform
distribution on the set of such pairs. If we want to select a
single pair, it is therefore reasonable to select the mean value
of $(\ell,h)$ under this distribution. It is known --- see, e.g.,
\cite{Ahsanullah2013,Arnold2008,David2003,Kosheleva2016,Kosheleva2016a,Lorkowski2014,Lorkowski2014a,Lorkowski2015,Lorkowski2016}
--- that this mean corresponds to $\ell=1/3$ and $h=2/3$. So, low
correlation corresponds to 1/3.

This explains why in both cases, we get good predictability for
1/3 of the students. \medskip

\noindent{\bf Why 1/2: analysis of the problem and the resulting
justification.} In the previous subsection, we argued that it is
reasonable to interpret low correlation between, e.g., the GPA and
the grade for a class as correlation of 1/3. Similarly, we can
argue that since the correlation between the GPA and the average
of the first grades is also low, we should also describe it by a
numerical value of 1/3.

We now have two events: \begin{itemz} \item predictability by
GPA (we will denote it by $A$) and \item predictability by the
average of the first few grades in a class (we will denote it by
$B$). \end{itemz} We know that the probability $P(A)$ of $A$ is
$p=1/3$, the probability $P(B)$ of $B$ is also 1/3, and the
correlation $r$ between $A$ and $B$ is also equal to 1/3. What is
then the probability $P(A\vee B)$ that for a random selected
student, his/her final grade in the class can be predicted based
either on the student's GPA or on the student's average grade for
the few class assignments?

By definition, the correlation $r$ between two random variables
$x$ and $y$ is equal to $$r=\frac{E[x\cdot y]-E[x]\cdot
E[y]}{\sigma[x]\cdot \sigma[y]},$$ where $E[\cdot]$ means the mean
value and $\sigma[\cdot]$ means standard deviation; see, e.g.,
\cite{Sheskin2011}.

In our case:\begin{itemz} \item $x$ is a random variable which
is equal to 1 if $A$ holds and 0 otherwise, and \item $y$ is a
random variable which is equal to 1 if $B$ holds, 0 otherwise.
\end{itemz}
Since each of the events $A$ and $B$ has probability $p=1/3$, we
can conclude that $E[x]=E[y]=p$. Here, the product of $x$ and $y$
is different from 0 only if both values $x$ and $y$ are different
from 0, i.e., when both events $A$ and $B$ occurred, so $E[x\cdot
y]=P(A\,\&\,B)$.

Here,
$$\sigma^2[x]=E[(x-E[x])^2]=p\cdot (1-p)^2+(1-p)\cdot p^2=p\cdot
(1-p)=\frac{1}{3}\cdot\frac{2}{3}=\frac{2}{9}.$$ Thus,
$\sigma[x]=\sigma[y]=\sqrt{2/9}$. Thus, the above formula for $r$
takes the form
$$\frac{1}{3}=\frac{P(A\,\&\,B)-1/9}{\sqrt{2/9}\cdot \sqrt{2/9}}=
\frac{P(A\,\&\,B)-1/9}{2/9}.$$ Thus,
$$P(A\,\&\,B)=\frac{1}{3}\cdot \frac{2}{9}+\frac{1}{9}=\frac{2}{27}+\frac{3}{27}=
\frac{5}{27}.$$ By additivity, we always have
$$P(A\vee B)=P(A)+P(B)-P(A\,\&\,B),$$ thus
$$P(A\vee B)=2\cdot
\frac{1}{3}-\frac{5}{27}=\frac{18}{27}-\frac{5}{27}=\frac{13}{27}.$$
This is very close to the empirical value of 1/2. Thus, we have
indeed provided a theoretical justification for this empirical
value.

\section{Towards Resulting Recommendations}

\noindent{\bf Before the class starts: analysis of the problem.}
Before the class starts, we need to check, for each student,
whether he/she belongs to the one-third of predictable by GPA. To
decide on this, we need to check, e.g., on the standard deviation
of all the student's previous grades.

Usually, in the US, we use letter grades. Specifically, the
original 0 to 100 numerical grade is transformed into one of the
few letter grades: A for excellent, B for good, C for
satisfactory, D for probably passing, and F for failing. In
computing the GPA, these grades are assigned the following
numerical values: A is 4, B is 3, C is 2, D is 1, and F is 0. Of
course, if we only use these 0 to 4 grades, we lose a lot of
information, so it is better to get and use the original 0-to-100
grades.

For each student, based on his/her previous class grades, we will
find the standard deviation. \begin{itemz} \item If this
standard deviation is low, then we can safely predict the
student's grade in the course based on his/her GPA.\item
Otherwise, we cannot make this prediction. \end{itemz} What
threshold should we use? Since about one third is predictable this
way, a natural idea is to sort all these standard deviations, and
consider those in the lower third predictable-by-GPA.

Thus, we arrive at the following recommendation.
\medskip

\noindent{\bf Before the class starts: resulting recommendation.}
In the beginning, when we do not know anything about the students,
any of them can be at-risk.

Before the class starts, to decrease the number of potential
at-risk students, we collect, for each student, his/her 0-to-100
grade in all previous classes. Based on these grades, we compute
the mean grade and the standard deviation corresponding to this
student.

We then sort all the standard deviations, and consider students
from the lower third. Those of these lower-third students whose
mean grade in previous classes is C or better are clearly not
at-risk, so they should be excluded from the list of possible
at-risk students.
\medskip

\noindent{\bf After the first few grades: analysis of the
problem.} First, we need to check, for each student from the
class, whether this student belongs to the one-third of
predictable-by-first-grades To decide on this, we need to check,
e.g., whether there is a correlation in the previous classes
between the average grade on the first several assignments in each
class and the final grade for the same class.

For this purpose, we need to know not only the student's 0-to-100
final grades in all previous classes, we also need to know, for
each of these previous classes, the student's average of the first
0-to-100 grades in this class.

Based on this information, we can compute, for each student, the
correlation between the average of the first few grades and the
final grade for the corresponding class. \begin{itemz} \item If
this standard correlation is high, then we can safely predict the
student's final grade in the course based on his/her average grade
from the first few assignments.\item Otherwise, we cannot make
this prediction. \end{itemz}
What threshold should we use? Since
about one third is predictable this way, a natural idea is to sort
all these correlations, and consider those in the upper third
predictable-by-first-grades.

Thus, we arrive at the following recommendation.
\medskip

\noindent{\bf After the first few grades: resulting
recommendation.} Before the class starts, for each student and for
each class that this student attended, we collect, in addition to
the student's 0-to-100 final grade for this class, an average of
the 0-to-100 grades for the first several assignments. Based on
this information, we compute, for each student, the correlation
between the final grade and the average-of-first grades.

We then sort all the resulting correlations, and consider students
from the upper third. Those of these upper-third students whose
average on the first few grades in this class is C or better are
clearly not at-risk, so they should be excluded from the list of
possible at-risk students.

The resulting list --- consisting of slightly more than half
students --- contains everyone who can potentially be at risk.

\section*{Acknowledgments}

This work was partially supported by the US National Science
Foundation via grant HRD-1242122 (Cyber-ShARE Center of
Excellence).

\begin{thebibliography}{99}

\bibitem{Ahsanullah2013}
 Ahsanullah M.,  Nevzorov V.B. and  Shakil M. An Introduction
to Order Statistics. Atlantis Press, Paris, 2013.

\bibitem{Arnold2008}
 Arnold B.C.,  Balakrishnan N. and  Nagaraja H.N. A First
Course in Order Statistics. Society of Industrial and Applied
Mathematics (SIAM), Philadelphia, Pennsylvania, 2008.

\bibitem{David2003}
 David H.A. and  Nagaraja H.N. Order Statistics. Wiley, New
York, 2003.

\bibitem{Kosheleva2016}  Kosheleva O.,  Kreinovich V.,  Lorkowski J. and
 Osegueda M. How to transform partial order between degrees into
numerical values. Proceedings of the 2016 IEEE
International Conferences on Systems, Man, and Cybernetics
SMC'2016, Budapest, Hungary, October 9--12, 2016.

\bibitem{Kosheleva2016a}
 Kosheleva O.,  Kreinovich V.,  Osegueda Escobar M. and
   Kato K. Towards the most robust way of assigning numerical
  degrees to ordered labels, with possible applications to dark
  matter and dark energy. Proceedings of the 2016 Annual Conference
  of the North American Fuzzy Information Processing Society
  NAFIPS'2016, El Paso, Texas, October 31 -- November 4, 2016.

\bibitem{Lorkowski2014} 
 Lorkowski J. and  Kreinovich V.
Interval and symmetry approaches to uncertainty --- pioneered by
Wiener --- help explain seemingly irrational human behavior: a case
study. Proceedings of the 2014 Annual Conference of the
North American Fuzzy Information Processing Society NAFIPS'2014,
Boston, Massachusetts, June 24--26, 2014.

\bibitem{Lorkowski2014a}  Lorkowski J. and  Kreinovich V.
Likert-type fuzzy uncertainty from a traditional decision making
viewpoint: how symmetry helps explain human decision making
(including seemingly irrational behavior. Applied and
Computational Mathematics, 2014, vol.~13, no.~3, pp.~275--298.

\bibitem{Lorkowski2015} 
 Lorkowski J. and  Kreinovich V.
Granularity helps explain seemingly irrational features of human
decision making. In: W. Pedrycz and S.-M.~Chen (eds.), 
Granular Computing and Decision-Making: Interactive and Iterative
Approaches, Springer Verlag, Cham, Switzerland, 2015, pp.~1--31.

\bibitem{Lorkowski2016}  Lorkowski J. and  Kreinovich V. Fuzzy logic ideas can
help in explaining Kahneman and Tversky's empirical decision
weights. In: L. Zadeh et al. (Eds.), Recent Developments
and New Direction in Soft-Computing Foundations and Applications,
Springer Verlag, to appear.

\bibitem{Rodriguez2017}  Rodriguez E. and  Chang G. Expected
academic performance in a lower level undergraduate structural
course. In: A. I\c{s}man and A. Eskiculami (eds.), 
Proceedings of the Joint 2017 International Conference on New
Horizons in Education (INTE), International Distance Education
Conference (IDEC), and International Trends and Issues in
Communication \& Media Conference (ITICAM), Berlin, Germany, July
17--19, 2017, pp.~1323--1331.

\bibitem{Sheskin2011}  Sheskin D.J. Handbook of Parametric and Nonparametric
Statistical Procedures. Chapman and Hall/CRC, Boca Raton,
Florida, 2011.

\end{thebibliography}

\maketitleEng
\endarticle