\startarticle 
\shrtitle{Secure Multi-Agent Quantum...}
\shrauthor{O.~Kosheleva, V.~Kreinovich}{O.~Kosheleva, V.~Kreinovich} 
\udc{004.056} 
\selectlanguage{english}
\date{02.12.2018}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Russian. If not compiled, it replaces the Russian text to English %%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Безопасная многоагентная квантовая коммуникация: на~пути к~наиболее эффективной схеме (педагогическое~замечание)} 
\author{О. Кошелева}{к.ф.-м.н., доцент}{olgak@utep.edu}
\author{В. Крейнович}{к.ф.-м.н., профессор}{vladik@utep.edu}
\affil{Техасский университет в Эль Пасо, США} 
\abstract{Во многих экономических и финансовых приложениях важно иметь безопасную связь. 
В настоящее время безопасность связи обеспечивается главным образом RSA-кодирование, 
но появляющиеся квантовые вычисления могут взломать этот код, делая его небезопасным. 
Один из способов сделать связь абсолютно безопасной --- использовать квантовое шифрование. 
Существующие схемы квантового шифрования нацелены на связь от агента к агенту; однако на практике 
нам часто требуется безопасная многоагентная связь, где каждый из агентов может безопасно 
отправлять сообщения всем остальным. В принципе, мы можем повторить схему агент-агент для 
каждой пары агентов, но для этого требуется большое количество сложных предварительных квантовых связей. 
В этой статье мы покажем, как минимизировать количество таких предварительных сообщений 
--- без ущерба для надёжности схемы всех пар.}
\keywords{обобщённые тригонометрические суммы, обработка сигналов}

\titleEng{Secure Multi-Agent Quantum Communication: Towards the
Most Efficient Scheme (a~Pedagogical~Remark)}     
\authorEng{Olga Kosheleva}{Ph.D. (Phys.-Math.), Associate Professor}{olgak@utep.edu}
\authorEng{Vladik Kreinovich}{Ph.D. (Phys.-Math.), Professor}{vladik@utep.edu}
\affilEng{University of Texas at El Paso, El Paso, Texas 79968, USA} 
\abstractEng{In many economic and financial applications, it
is important to have secure communications. At present,
communication security is provided mostly by RSA coding, but the
emergent quantum computing can break this encoding, thus making it
not secure. One way to make communications absolutely secure is to
use quantum encryption. The existing schemes for quantum
encryption are aimed at agent-to-agent communications; however, in
practice, we often need secure multi-agent communications, where
each of the agents has the ability to securely send messages to
everyone else. In principle, we can repeat the agent-to-agent
scheme for each pair of agents, but this requires a large number
of complex preliminary quantum communications. In this paper, we
show how to minimize the number of such preliminary communications
--- without sacrificing reliability of the all-pairs scheme.}
\keywordsEng{Quantum communication, multi-agent communication}

\maketitle

\section{Need for Secure Group Communications}

\noindent{\bf Need for secure communications.} In e-commerce and
e-finance, it is important to preserve privacy and confidentiality
of all the transactions. In other words, we need to make sure that
e-commerce and e-finance are based on a secure communication
scheme.
\medskip

\noindent{\bf How communications are made secure now.} At present,
secure communications are based mostly on the RSA algorithm; see,
e.g., \cite{Cormen2009}. In this scheme, the communicator $A$
selects two large prime numbers $P_1$ and $P_2$ that he/she keeps
private, and releases their product $P=P_1\cdot P_2$ into the
public domain. This public code $P$ can then be used any anyone to
encode the messages they send to $A$. To decode the messages, one
needs to know the factors $P_1$ and $P_2$. Since $A$ knows these
factors, $A$ can decode these messages.

The security of this encoding scheme is provided by the fact that
no efficient algorithm is known for factoring large integers --
and RSA algorithms use 100-digit and longer factors $P_i$. In
principle, we can factor an integer $P$ by trying all possible
prime numbers $p\le \sqrt{P}$. This works for small $P$, but for a
number with 100 decimal digits, testing all prime numbers $\le
\sqrt{P}$ requires $\sqrt{10^{100}}=10^{50}$ computational steps
--- which make this procedure much longer than the lifetime of the
Universe.
\medskip

\noindent{\bf Why we need quantum communications.} For factoring
large integers on the usual computers, no efficient algorithm is
known. However, it is known that if we consider quantum computers,
then factoring large integers becomes feasible; see, e.g.,
\cite{Nielsen2000,Shor1994,Shor1997,Williams2000}. At present, we
do not yet have quantum computers powerful enough to decode the
usual RSA message, but engineers are designing more and more
powerful quantum computers, and sooner or later RSA-encoded
communications will no longer be secure.

Good news is that the same quantum physics that makes RSA not
secure also provides us with a secure way to communicate, known as
quantum cryptography; see, e.g.,
\cite{Galindo2018,Nielsen2000,Williams2000}. In contrast to
quantum computing, which is mostly the thing of the future,
quantum cryptography is a practical scheme, it has been used for
decades already.

In this scheme --- it is described in some detail in the appendix
--- two agents that want to communicate in the future exchange
quantum signals. By analyzing these signals, they come up with a
sequence $s=s_1s_2\ldots s_n$ of bits (0s and 1s) $s_1$, $s_2$,
\ldots, $s_n$ which they both know but which is not known to
anyone else.

This sequence of 0s and 1s can then be used as {\it one-time pad}.
Namely, if one of the two agents needs to send a message
$m=m_1m_2\ldots m_n$ consisting of bits $m_1,\ldots,m_n$, then:
\begin{itemize}
\item the sender computes and send the encoded signal $e=s\oplus
m$, where $$(s\oplus m)_i\stackrel{\rm def}{=}s_i\oplus m_i$$ and
$\oplus$ means addition modulo 2 --- which differs by the usual
addition of bits only when $s_i=m_i=1$, in which case $1\oplus
1=0$; \item the receiver, after getting the signal, reconstructs
the original message as $m=s\oplus e$.
\end{itemize}
Indeed, for addition modulo 2, we always have $a\oplus a=0$, hence
$$s\oplus e=s\oplus (s\oplus m)=(s\oplus s)\oplus m=0\oplus m=m.$$

Once the message is sent, the pair of agents again perform
preliminary quantum communications and generate a new one-time
pad, etc.
\medskip

\noindent{\bf Need for multi-agent communications.} In economic
and financial applications, it is often important to have
multi-agent communications. Such communications are especially
important for decentralized schemes (like blockchain-based
schemes), where each record of financial transactions is stored in
many different locations.

In such a scheme, we have several agents. Let us denote the
overall number of agents who need to communicate with each other
by $N$. The system should be ready for each of these agents
$A_1,\ldots,A_N$ to send communication to everyone else.
\medskip

\noindent{\bf Using pairwise communication scheme for multi-agent
communication: a straightforward idea.} In principle, we should be
ready for communications between each pair of agents. For $N$
agents, there are $\displaystyle\frac{N\cdot (N-1)}{2}$ such
pairs. So, a straightforward idea is to repeat the
quantum-communication protocol for each pair $(A_j,A_k)$ and thus
get $\displaystyle\frac{N\cdot (N-1)}{2}$ one-time pads
$s^{(j,k)}$ corresponding to these pairs.

If agent $A_j$ needs to send a message to all the other agents
$A_1$, $A_2$, \ldots $A_{j-1}$, $A_{j+1}$, \ldots, $A_n$, this
agent will use a one-time pad $s^{(j,k)}$ to communicate with the
$k$-th agent.
\medskip

\noindent{\bf Can we do it more efficiently?} The above scheme
requires $\displaystyle\frac{N\cdot (N-1)}{2}$ preliminary quantum
communications --- and preliminary quantum communications are the
most complex part of the general quantum communication protocol.
Can we have fewer preliminary quantum communications?
\medskip

\noindent{\bf We can do it, but at the expense of reliability.}
One possibility to have a more efficient multi-communication
scheme is to select one agent as a hub and set up one-time pads
between this selected agent and everyone else. This way, we have a
secure communication channel between the hub agent and every other
agent.

Then, if an agent $A_j$ wants to transmit a message to everyone
else, this agent first sends this message to the hub agent, and
then the hub agent sends it to everyone else.

This scheme is efficient --- it needs only $N-1$ preliminary
quantum communications --- but it is not as reliable as the
original scheme: indeed, if the hub agent is not functioning well
(which happens), then the original scheme still works while the
hub scheme does not. So, we arrive at the following question.
\medskip

\noindent{\bf Can we have an efficient scheme without decreasing
reliability: formulation of the problem and our answer.} Can we
have an efficient scheme without decreasing reliability?

In this paper, we show that this is indeed possible: namely, we
provide a multi-agent communication scheme which is maximally
efficient and at the same time as reliable as the all-pairs
scheme.

\section{Towards the Optimal Multi-Agent Quantum Communication
Protocol}

\noindent{\bf Lower bound on the number of preliminary quantum
communications.} The only way to have secure communications is to
have a secure one-time pad provided by preliminary quantum
communications. A pair of agents may get a one-time pad either
directly from the mutual preliminary quantum communication, or by
somehow combining one-time pads provided by other pairs.

Each agent can thus securely communicate only with agents which
are either directly connected with this agent by preliminary
quantum communications, or connected by a chain in which each
agent is connected to the next one by preliminary quantum
communications between them.

Since we want each agent to be able to securely communicate with
any other agent, every two agents must be connected by such a
chain. We start with a single agent. At each point, we have a set
of agents that can be thus connected with the agent $A_1$. Each
new preliminary quantum agent-to-agent communication adds may add
one agent to this list --- if it connects this new agent with one
of the agents which are already on this list. Thus, each new
comunication adds no more than one new agent to this list. We
start with a single agent, so after performing $k$ agent-to-agent
preliminary quantum communications, we add no more than $k$ agents
to the original list, and thus, we have $\le k+1$ agents connected
to $A_1$. We want to have all $N$ agents connected to $A_1$, so we
must have $N\le k+1$ and thus, $k\ge N-1$.

This is the desired lower bound: to enable each agent to securely
communicate with every other agent, we need to perform at least
$N-1$ agent-to-agent preliminary quantum communications.
\medskip

\noindent{\bf Towards a scheme that implements this lower bound.}
Let us show that it is possible to have a secure communication
protocol that requires exactly $N-1$ agent-to-agent preliminary
quantum communications. Due to what we have just shown, this
algorithm is the most efficient one --- in the sense that it
requires the smallest possible number of agent-to-agent
preliminary quantum communications.

Let one of the agents --- let us denote this agent by $A_1$ ---
perform the preliminary step of quantum communication with every
other agent $A_1$, \ldots, $A_m$. As a result, for each
$j=2,3,\ldots,N$, both $A_1$ and $A_j$ know a one-time pad
$s^{(j)}$ --- which no one else knows.

Then, $A_1$ uses a random number generator to generate a random
string $s$ and then sends by an open channel, to each agent $A_j$,
a string $s\oplus s^{(j)}$. Each agent can reconstruct $s$ as
$(s\oplus s^{(j)})\oplus s^{(j)}$. Thus, all $N$ agents now have
the same one-time pad $s$ that they can use for the future
multi-agent communication.

\section*{Acknowledgments}

This work was partially supported by the US National Science
Foundation via grant HRD-1242122 (Cyber-ShARE Center of
Excellence).

\begin{thebibliography}{9}

\bibitem{Cormen2009} Th. H. Cormen,
C. E. Leiserson, R. L. Rivest, and C. Stein, {\it Introduction to
Algorithms}, MIT Press, Cambridge, Massachusetts, 2009.

\bibitem{Feynman2005}
R. Feynman, R. Leighton, and M. Sands, {\it The Feynman Lectures
on Physics}, Addison Wesley, Boston, Massachusetts, 2005.

\bibitem{Galindo2018} O. Galindo, V. Kreinovich, and O. Kosheleva,
``Current
quantum cryptography algorithm is optimal: a proof'', {\it
Proceedings of the IEEE Symposium on Computational Intelligence
for Engineering Solutions CIES'2018}, Bengaluru, India, November
18--21, 2018.

\bibitem{Nielsen2000} M. Nielsen and I. Chuang, {\it Quantum Computation and
Quantum Information}, Cambridge University Press, Cambridge, 2000.

\bibitem{Shor1994} P. Shor, ``Polynomial-time algorithms for prime
factorization and discrete logarithms on a quantum computer'',
{\it Proceedings of the 35th Annual Symposium on Foundations of
Computer Science}, Santa Fe, New Mexico, November 20--22, 1994.

\bibitem{Shor1997} P. Shor, ``Polynomial-time algorithms for prime
factorization and discrete logarithms on a quantum computer'',
{\it SIAM J. Sci. Statist. Comput.}, 1997, Vol. 26, pp.~1484-ff.

\bibitem{Thorne2017} K. S. Thorne and R. D. Blandford,
{\it Modern Classical Physics: Optics, Fluids, Plasmas,
Elasticity, Relativity, and Statistical Physics}, Princeton
University Press, Princeton, New Jersey, 2017

\bibitem{Williams2000} C. P. Williams and S. H. Clearwater, {\it
Ultimate Zero and One}, Copernicus, New York, 2000.

\end{thebibliography}

%\appendix
\section*{A. Quantum Communication: A Brief Reminder}

\noindent{\bf Quantum background: a brief reminder.} In quantum
mechanics (see, e.g., \cite{Feynman2005,Thorne2017}) each bit, in
addition to the usual two states 0 and 1 (which are denoted by
$|0\rangle$ and $|1\rangle$), it is also possible to have a
superposition $c_0\cdot |0\rangle +c_1\cdot |1\rangle$, where
$c_i$ are complex numbers for which $|c_0|^2+|c_1|^2=1$. If in
this state, we try to find out whether we have 0 or 1, we get 0
with probability $|c_0|^2$ and 1 with probability $|c_1|^2$.

In particular, the standard quantum communication algorithm uses
the following two states:
$$|0'\rangle\stackrel{\rm def}{=}\frac{1}{\sqrt{2}}\cdot |0\rangle
+\frac{1}{\sqrt{2}}\cdot |1\rangle$$ and
$$|1'\rangle\stackrel{\rm def}{=}\frac{1}{\sqrt{2}}\cdot |0\rangle
-\frac{1}{\sqrt{2}}\cdot |1\rangle.$$

To reconstruct the state, we need to use a measuring instrument. A
measuring instrument can be tuned:\begin{itemize} \item either to
detect the usual states $|0\rangle$ and $|1\rangle$ (we will
denote this by +) \item or to detect the quantum states
$|0'\rangle$ and $|1'\rangle$ (we will denote this tuning by
$\times$).
\end{itemize}
In this case:
\begin{itemize}
\item If the measuring instrument's tuning matches the signal --
e.g., if the signal is $|0\rangle$ or $|1\rangle$ and the
instrument is tuned on $|0\rangle$ or $|1\rangle$ --- then the
instrument reconstructs the original signal. \item On the other
hand, if there is a mismatch between the instrument's tuning and
the signal --- e.g., if the signal is $|0'\rangle$ or $|1'\rangle$
while the instrument is tuned on $|0\rangle$ or $|1\rangle$ --
then, irrespective of the signal, the instrument returns 0 or 1
equal with probability 1/2, and the original signal is lost.
\end{itemize}
\medskip

\noindent{\bf The actual quantum communication scheme.} The
sending agent $A$ runs, several times, a random number generator
that generates 0 or 1 with equal probability 1/2. As a result, we
get a multi-bit sequence $r_1\ldots r_c$. Then, $A$ runs the same
random number generator $n$ more times, generating $c$ more bits
$t_1\ldots t_c$. For each $i$: \begin{itemize} \item if $t_i=0$,
then $A$ uses the + tuning to send the signal $t_i$, i.e., sends
$|0\rangle$ if $r_i=0$ and sends $|1\rangle$ if $r_i=1$; \item if
$t_i=0$, then $A$ uses the $\times$ tuning to send the signal
$t_i$, i.e., sends $|0'\rangle$ if $r_i=0$ and sends $|1'\rangle$
if $r_i=1$.
\end{itemize}
The receiving agent $B$ also runs its own random random number
generator, generating yet another sequence $b_1,\ldots,b_c$ of $c$
bits. Then, for each $i$:
\begin{itemize}
\item if $b_i=0$, then $B$ uses the + tuning to measure the
received signal, and \item if $b_i=1$, then $B$ uses the $\times$
tuning to measure the received signal.\end{itemize}

For those bits for which there is a match between the signal and
the tuning, i.e., for which $t_i=b_i$, $B$ gets exactly the
original signal $r_i$. For every other index $i$, the result of
$B$'s measurements is 0 or 1 with probability 1/2 --- and the
original signal $r_i$ is lost.

Now, $A$ openly sends, to $B$, all the bits $t_i$. In half of the
cases, $b_i$ coincides with $t_i$. The agent $B$ sends, to $A$,
the list of such $i$'s. For these $i$'s --- and there are $c/2$
such $i$'s --- both agents know the value $r_i$. The sequence of
all these common values is the desired one-time pad of length
$n\approx c/2$ that $A$ and $B$ can now use to securely
communicate.
\medskip

\noindent{\bf Why is this scheme secure.} The security of this
scheme is based on the fact that an eavesdropped does not know
which tuning was used for each bit $i$. If the eavesdropper tries
to measure the signal, it will thus, in half of the cases, use the
wrong tuning, and thus, the original signal will be lost --- i.e.,
replaced with a random bit.

As a result, even for some of the bits for which $t_i=b_i$, due to
this replacement, the bits measured by $B$ will be, in general,
different from $r_i$. To detect eavesdropping, $A$ also sends to
$B$, by open channel, some of the bits $r_i$ corresponding to the
cases when $t_i=b_i$. \begin{itemize} \item If for some $i$, $B$
measured a bit different from $r_i$, this means that there was an
eavesdropper. \item If all the test-bits $r_i$ are reproduced by
$B$ exactly --- this means that the channel was secure, there was
no eavesdropping.
\end{itemize}

\clearpage
\maketitleEng
\endarticle